Fluent statistical computing interfaces for biological data analysis

### 2020-02-23 Rmd/01-intro.Rmd
delete abstract; rewrite intro

overview

describe main topic - research questions
chapters 1, 2, 3, 4 (your contributions to the joint work)
scope

>>>>

Exploratory data analysis (EDA) is a vital element of the modern 
statistical workflow - it is an analyst’s first pass at 
understanding their data; revealing all it’s messes and (possibly) 
uncovering hidden insights [@Tukey1977-pn; @Grolemund2017-uc]. 
It is an iterative process invovling both computation and visualization 
that can generate new hypotheses that can be tested and formalised using 
stastical modelling. Moreover, as datasets grow in complexity and become
increasingly heterogenous and multidimensional, 
the use of EDA becomes vital to ensure the integrity and quality of analysis 
outputs.  This is certainly true in high-throughput biological 
data analysis, where there are constraints on computation time and 
memory, leading to the situation that as data becomes larger,
EDA becomes increasingly difficult.

The approach taken by the suite of software packages collectively known as the *tidyverse* 
is an attempt to formalise aspects of the EDA process in the R programming language
under a single semantic known as *tidy data* [@baser; @tidyverse; @Wickham2014-jc]. 
Simply put, a *tidy data* set is a rectangular table where each row of the table corresponds to an 
observation and each column corresponds to a variable. There is a surprisingly
large amount of utility that can be achieved with this definition. By having
each column representing a variable,  variables can be mapped
to graphical aesthetics  of visualizations. This enables the grammar of graphics as implemented
by *ggplot2* [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces
as implemented by *tidyverse* and in particular the *dplyr* package are *fluent*; 
they form a domain specific language (DSL) that gives users a mental model 
for performing and composing common data transformation tasks 
[@Wickham2017-dplyr; @FowlerFluent].

It is unclear whether the *fluent* interfaces as implemented using the
*tidy data* framework can be more generally applied and useful
in fields such as high-throughput biology where domain specific semantics as
implemented through more complex data structures are required. This is particularly
true in the Bioconductor ecosystem, where much thought has gone into the
design of data structures that enable interoperability between different tools 
[@Huber2015-ei]. One contribution of this thesis is the implementation
of a *fluent* interface for genomics data that is expressive enough to
enable EDA and be used as a building block for more software.

## A grammar for genomic data analysis

## Integration and representation of genomic data structures

## Visual analytics for dimension reduction and clustering workflows

## Summary

This thesis presents a set of tools for performing exploratory 
data analysis in the context of genomics and transcriptomics.  The first tool, “plyranges” provides a domain specific language for 
reasoning about range based genomics data that simplifies common 
operations such as overlaps and aggregations.  Next I extend 
“plyranges”  to develop “superintronic” which presents an 
approach for discovering interesting genomic regions within an 
experimental design. We apply this tool to explore intron signal 
in transcriptomic data. Finally, I turn my attention to focus on 
embedding methods currently used for visualisation of large data 
sets…
